Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd

Our fourth goal was to extract higher-order dimensions of leader qualities and functions from our researcher coded variables. We used several methods to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. Logistic PCA generalizes Pearson's original formulation of PCA to find a projection of the natural parameters of the saturated model that minimizes the Bernoulli deviance [@logisticPCA].

@@

Our fifth goal was to assess variation in our "feature" dimensions by group context, subsistence strategy, continental region, and leader sex.

@@

## Shaman leaders

Shamans and other leaders with supernatural qualities were surprisingly common in the ethnographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of shamanism as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure \@ref(fig:shamanism).

@@

We identify these seven named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. Because features were aggregated variables, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

@@

In our view, the dominance-prestige and "dual-model" theories of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "managers". These models, in short, have conflated the ability to provide "benefits" with knowledge and expertise, and the ability to impose "costs" with physical formidability. @van_vugt_dual_2019, for instance, state that the source of deference for prestige-style leadership is "Information asymmetry", and for dominance-style, "Power asymmetry" (their Table 1). Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability. The use of knowledge to impose costs is widely recognized by many social scientists. The role of gossip and other forms of indirect aggression in social dominance and achieving leadership roles, for instance, has been extensively investigated in WEIRD populations [e.g., @hawley_ontogenesis_1999; @hess_gossip_2017; @hess_sex_2006].

We therefore tenatively advance an alternative scheme, similar to @von_rueden_roots_2014[] [and inspired by @kaplan_embodied_2003; @kaplan_evolutionary_2009; @borgerhoff_mulder_intergenerational_2009], whereby individuals possess some combination of cognitive capital, social capital, and somatic capital that they use to provide benefits and impose costs. If their abilities to provide benefits or impose costs are exceptional relative to other group members, they gain status via biological market mechanisms [@hammerstein_biological_2016; @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2019]. Hence, we should be able to find examples of leaders with various combinations of these five elements.^[We omit material capital because this could be both a means to leadership as well as a benefit of leadership.]

```{r fivefold, fig.width=10, fig.cap="The intersections of evidence for our five basic elements of leadership among all text records (an 'upset' plot) [@gehlenborg2019]. Horizontal bars (left): The total number of text records in each element. Dots (bottom): Elements that both have evidence (intersection). Vertical bars (top): The number of text records in that intersection (with lower limit set at 3). Red: The number of text records involving shamans or leaders with supernatural qualities. "}

upset(
  df_fivefold2,
  nintersects = 18,
  order.by = 'freq',
  point.size = 5, 
  text.scale = 2,
  queries = list(list(query = shamanfn, color = 'red', active=T))
)

As Figure \@ref(fig:fivefold) reveals, descriptions of leadership in the ethnographic record includes a heterogeneous mix of these traits. In particular, not all leaders have evidence for high *Social capital*, leaders with *Cognitive capital* provide benefits *and* impose costs, and *Somatic capital* is rarely mentioned, but when it is, it is associated with providing benefits as well as imposing costs. Shamans are associated with evidence for *Cognitive capital* and imposing costs.

>>>>

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. Logistic PCA generalizes Pearson's original formulation of PCA to find a projection of the natural parameters of the saturated model that minimizes the Bernoulli deviance [@logisticPCA].

@@

Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex.

Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories.

@@

@@

We identify these seven named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership reviewed in the introduction, with the possible exception of competencies, a single feature that included at least two elements (prestige and dominance). Because features were aggregated variables, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

## Shaman leaders and competencies

The competencies feature combined dimensions of prestige-style leadership, such as knowledge and experience, with dimensions of dominance-style leadership, such as aggressiveness and feared. It also included supernatural qualities. Shamans and other leaders with supernatural qualities were surprisingly common in the ethnographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of shamanism as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure \@ref(fig:shamanism).

## Cognitive, social, and somatic capital

Motivated by our feature analysis, and the apparent importance of shaman leaders, we tenatively advance a scheme, similar to @von_rueden_roots_2014[] [and inspired by @kaplan_embodied_2003; @kaplan_evolutionary_2009; @borgerhoff_mulder_intergenerational_2009], whereby leaders possess some combination of cognitive capital, social capital, and somatic capital that they deploy to provide benefits and impose costs (for further justification, see the Discussion section). Hence, we should be able to find examples of leaders with various combinations of these five elements.^[We omit material capital because this could be both a means to leadership as well as a benefit of leadership.]

```{r fivefold, fig.width=10, fig.cap="The intersections of evidence for our five basic elements of leadership among all text records (an 'upset' plot) [@gehlenborg2019]. Horizontal bars (left): The total number of text records in each element. Black dots: has evidence. Gray dots: does not have evidence. Vertical bars (top): The number of text records in that intersection (with lower limit set at 3). Red: The number of text records involving shamans or leaders with supernatural qualities. "}

p <- upset(
  df_fivefold2,
  nintersects = 18,
  order.by = 'freq',
  point.size = 5, 
  text.scale = 2,
  queries = list(list(query = shamanfn, color = 'red', active=T))
)
pdf("Figures/upset.pdf", width = 12, height = 9)
p
dev.off()
p
```

Figure \@ref(fig:fivefold) reveals that descriptions of leadership in the ethnographic record includes a heterogeneous mix of these five elements. In particular, not all leaders have evidence for high *Social capital*, leaders with *Cognitive capital* provide benefits *and* impose costs, and *Somatic capital* is rarely mentioned, but when it is, it is associated with providing benefits as well as imposing costs. Shamans are associated with evidence for *Cognitive capital* and imposing costs.

@@

In our view, the dominance-prestige and "dual-model" theories of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "managers". These models, in short, have conflated the ability to provide "benefits" with knowledge and expertise, and the ability to impose "costs" with physical formidability. @van_vugt_dual_2019, for instance, state that the source of deference for prestige-style leadership is "Information asymmetry", and for dominance-style, "Power asymmetry" (their Table 1). Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability. The use of knowledge to impose costs is widely recognized by many social scientists. The role of gossip and other forms of indirect aggression in social dominance and achieving leadership roles, for instance, has been extensively investigated in WEIRD populations [e.g., @hawley_ontogenesis_1999; @hess_gossip_2017; @hess_sex_2006] and also among hunter-gatherers [@hess_interpersonal_2010].

If their abilities to provide benefits or impose costs are exceptional relative to other group members, they gain status via biological market mechanisms [@hammerstein_biological_2016; @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2019].
